---
title: "The jazz of web"
subtitle: "A network jam for R (R study group)"
author: "Giulio Valentino Dalla Riva"
date: "June 17, 2016"
output: 
  ioslides_presentation: 
    highlight: espresso
    incremental: yes
    keep_md: yes
    smaller: yes
    widescreen: yes
---

## Webs

Relational structures.

- Food webs
- Proteome, Diseosome, -ome -ome -ome
- You name it
- Jazz musicians collaborations

## Webs

Today is [jazz](https://cordame.bandcamp.com/album/satie-variations)!

## R packages | What do we need?

```{r,echo=T,warning=FALSE,message=FALSE,error=FALSE}
library("igraph") # To handle the webs objects and plot them

options(stringsAsFactors = FALSE) # To handle the web datasets
```

## R packages | What may we need?

Useful but not today:

- `library("dplyr")` to filter, select, mutate
- `library("magrittr")` use pipes, everything is nicer ( `%>%` )
- ...

## Ingredients: interactions data | Better open

List of edges of the network of Jazz musicians. Data compiled by members of our group. Please cite P.Gleiser and L. Danon , Adv. Complex Syst.6, 565 (2003).
<http://deim.urv.cat/~alexandre.arenas/data/welcome.htm>

```{r}
Source.url <- "http://deim.urv.cat/~alexandre.arenas/data/xarxes/jazz.zip"
```

## Ingredients: interactions data | Better open

What do we have

```{r}
tempfile() -> temp.file # we create a temp file for the download,
Source.url %>% download.file(temp.file, mode="wb") # download it from the web
temp.file %>% unzip(list=TRUE) # and we use unzip to list what's in there
```

## Ingredients: interactions data | Better open

And what is that?

```{r}
temp.file %>% unzip('jazz.net') # we unzip the right file
'jazz.net' %>% # and then
  readLines %>% # we read as lines of text
  .[1:10] %>% # we select the first 10 lines
  cat(sep="\n") # and print them on a column
```

## Ingredients: interactions data | Better open

We like it: it seems a data frame of pair of nodes (and maybe weights?)

```{r}
'jazz.net' %>%
  read.table(skip = 3) -> # we read it as a table and skip the first 3 rows
edge.df # and store it in `edge.df`

temp.file %>% unlink # to keep things clean we unlink the temp file
```

## Ingredients| Knead the data to a web

Let's take a look at it

```{r}
edge.df[1:10,]
```

## Ingredients| Knead the data to a web

Wait, why three columns, which one is for the weights?

Let's see what they contain

```{r}
edge.df[,1] %>% unique %>% head
edge.df[,2] %>% unique %>% head
edge.df[,3] %>% unique %>% head
```

## Ingredients| Knead the data to a web

We can give some meaningful name to the variables

```{r}
names(edge.df) <- c("Source","Sink","Weight")
```

## Ingredients| Knead the data to a web

The igraph solo

```{r}
edge.df[,c("Source","Sink")] %>% # we select the right columns
  graph.data.frame(directed=FALSE) -> # call igraph (for the first time)
Jazz.graph # and obtain a graph (an igraph object)!
```

Memento: igraph goes `from`, `to` (or `source`, `sink`).

## Ingredients| A graph!

What do we have?

How many vertices in the graph? How many edges?
```{r}
Jazz.graph %>% vcount # count vertices
Jazz.graph %>% ecount # count edges
```

or simply
```{r}
Jazz.graph %>% summary
```

## Plot it!

What do we have?

```{r}
Jazz.graph %>% plot
```


## Plot it!

That was blah...

- Multiple links.
- Node position is not informative (we made Edward Tufte sad).
- We can do much better!

## What do we want to know?

Well, What about assortativity (homophyly):
do jazz musician with a lot of collaboration tend to collaborate more with other super collaborators or with not so collaborative musicians?

```{r}
Jazz.graph %>% # we need to
  simplify -> # remove duplicate edges!
Jazz.graph

Jazz.graph %>%
  assortativity_degree(directed = FALSE) -> # compute assortativity
Jazz.a
```

## And is that significant?

```{r}
Jazz.a
```

Is that a big, small, cute number?

We need to compare it with the expected value from our _null model_.

## Null model

A null model is as random as Jazz improvisation is.

> _True originality, and thus true creativity, never takes place in a historical vacuum; it is always rooted to something that has gone before._  
> **Brad Mehldau** - [Jazz's high stakes and tragic failures](http://www.bradmehldau.com/new-page)

## Null model: constraints and randomization

For example, let's keep fixed each node degree and change everything else:

```{r}
Jazz.graph %>%
  get.edgelist(name = FALSE) -> # we get the table of edges
edge_random.df

edge_random.df[,2] %>%
  sample -> # we shuffle all the sinks
edge_random.df[,2]

edge_random.df %>%
  graph.data.frame(directed=FALSE) -> # and we create a shuffled web
Jazz_random.graph
```

## Null model: constraints and randomization

We better wrap all that in a function:

```{r}
randomize_fixdegrees <- function(graph){
  library(igraph) # we need the igraph library
  edge_random.df <- get.edgelist(graph, name = FALSE)
  edge_random.df[,2] <- sample(edge_random.df[,2])
  random.graph  <-  graph.data.frame(edge_random.df, directed=FALSE)
  return(random.graph)
}
```

## Null model: constraints and randomization

What assortativity value has a random graph with the same degree sequence of `Jazz.graph`?

```{r}
Jazz.graph %>% # our observed graph
  randomize_fixdegrees %>% # we shuffle it
  assortativity_degree(directed = FALSE) # and compute its assortativity
```

## Null model: constraints and randomization

And let's wrap it in a function, is always better!

```{r}
random_assortativity <- function(graph,randomizer){
  Rand.graph <- randomizer(graph) # we shuffle our observed graph with our randomization function
  Rand.a <- assortativity_degree(Rand.graph, directed = FALSE) # we compute its assortativity
  return(Rand.a) # and we return it
}
```

## Simulated p-value fyeah! | Let's have a chat about p-values later

Let's run a simulation!

```{r}
N <- 420 # how many replications? 42 * 10!

Rand.a.vec <- replicate(N, # we replicate N times
              random_assortativity(Jazz.graph,randomize_fixdegrees)) # our sampling

# With which frequency the random graph had a higher assortativity?
p.sim <- sum(Rand.a.vec > Jazz.a)/N # let's name it p
p.sim
```


## Effect size plot

You should not just and only trust `p`. Let's, for example, look also the distribution of the assortativity in our null model simulation

```{r}
hist(Rand.a.vec)
abline(v = Jazz.a, col = "red", lwd = 3) # and see where the observed is
```


## Well done!

You made a discovery: jazz musician with a lot of collaboration tend to collaborate between themselves.